---
title: "Stat 437 HW1"
author: 
     - Samantha Gregoryk 
header-includes:
   - \usepackage{bbm}
   - \usepackage{amssymb}
   - \usepackage{amsmath}
   - \usepackage{graphicx,float}
   - \usepackage{natbib}
output:
  pdf_document: default
fontsize: 11pt
---

```{r, echo = TRUE, warning = FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

# import libraries
library(nycflights13)
library(dplyr)
library(ggplot2)
```

# Problem 1

Please refer to the NYC flight data `nycflights13` that has been discussed in the lecture notes and whose manual can be found at https://cran.r-project.org/web/packages/nycflights13/index.html. We will use `flights`, a tibble from `nycflights13`.

You are interested in looking into the average `arr_delay` for 6 different `month` 12, 1, 2, 6, 7 and 8, for 3 different `carrier` "UA", "AA" and "DL", and for `distance` that are greater than 700 miles, since you suspect that colder months and longer distances may result in longer average arrival delays. Note that you need to extract observations from `flights` and obtain the needed sample means for `arr_delay`, and that you are requird to use `dplyr` for this purpose.

The following tasks and questions are based on the extracted observations.

(1.a) In a single plot, create a density plot for `arr_delay` for each of the 6 months with `color` aesthetic designated by `month`. Note that you need to convert `month` into a factor in order to create the plot. What can you say about the average `arr_delay` across the 6 months?

```{r, 1a, echo = TRUE, warning = FALSE}
# average `arr_delay` for 6 different `month` 12, 1, 2, 6, 7 and 8, for 3 different `carrier` "UA", "AA" and "DL"
df = filter(flights, month %in% c(12, 1, 2, 6, 7, 8), 
       carrier %in% c('UA', 'AA', 'DL'), distance > 700)

# density plot

# factor month 
fac.month = as.factor(df$month)

# create density plot for arr_delay for each of the 6 months
# represent a density by dashed line
ggplot(df, aes(x = arr_delay, color = fac.month)) + geom_density(linetype = "dashed") + theme_bw()
```

1A: What can you say about the average `arr_delay` across the 6 months? \vspace{0.2in}
\newline The average delay for flights for the months of 12, 1, 2, 6, 7, 8 show that they are skewed to the left so the amount of delays for each month are minimal. As the year goes on, the average amount of delays decrease. 

\newpage (1.b) In a single plot, create a boxplot for `arr_delay` for each of the 3 carriers. What can you say about the average `arr_delay` for the 3 carriers?

```{r, 1b, echo = TRUE, warning = FALSE}
# boxplot
ggplot(df, aes(x = carrier, y = arr_delay)) + geom_boxplot() + theme_bw() + stat_summary(fun.y = mean, geom="point", shape = 23,size = 4)
```

1B: What can you say about the average `arr_delay` for the 3 carriers? \vspace{0.2in}
\newline All carriers have similar interquartile ranges around 0 average delays. AA and DL have similar outliers of average delays while UA has less average delays than the other carriers. 

\newpage (1.c) Create a pie chart for the 3 carriers where the percentages are the proportions of observations for each carrier and where percentages are superimposed on the sectors of the pie chart disc.

```{r, 1c, echo = TRUE, warning = FALSE}
pc = df %>% group_by(carrier) %>%
  count() %>% ungroup() %>%
  mutate(percentage=`n`/sum(`n`)) %>%
  arrange(desc(carrier))

# create labels using the percentages
pc$labels <- scales::percent(pc$percentage)

# pie chart
ggplot(pc) + geom_bar(aes(x = "", y = percentage, fill = carrier), stat = "identity", width = 1) + coord_polar("y", start = 0) + theme_void() + geom_text(aes(x = 1, y = cumsum(percentage) - percentage/2, label = labels))
```

\newpage (1.d) Plot `arr_delay` against `distance` with `facet_grid` designated by `month` and `carrier`.

```{r, 1d, echo = TRUE, warning = FALSE}
# ggplot
p1 = ggplot(df) + geom_point(mapping = aes(x = arr_delay, y = distance))

# facet_grid
p1 + facet_grid(month ~ carrier)
```

\newpage (1.e) For each feasible combination of values of `month` and `carrier`, compute the sample average of `arr_delay` and save them into the variable `mean_arr_delay`, and compute the sample average of `distance` and save these averages into the variable `mean_distance`. Plot `month` against `mean_arr_delay` with `shape` designated by `carrier` and `color` by `mean_distance`, and plot `month` against `mean_arr_delay` with `shape` designated by `carrier` and `color` by `mean_distance` and annotate each point by its associated `carrier` name. 

```{r, 1e, echo = TRUE, warning = FALSE}
# For each feasible combination of values of `month` and `carrier`
# compute the sample average of `arr_delay` and save them into the variable `mean_arr_delay`
# compute the sample average of `distance` and save these averages into the variable `mean_distance`

# change NA values for computing averages
df = as.data.frame(df)
df[is.na(df)] <- 0

# find the mean average of arr_delay for each carrier for each month

# filter carriers
cAA = df %>% filter(carrier %in% 'AA')
cUA = df %>% filter(carrier %in% 'UA')
cDL = df %>% filter(carrier %in% 'DL')

# filter months for carrier CAA
m12.CAA = cAA %>% filter(month %in% 12)
m1.CAA = cAA %>% filter(month %in% 1)
m2.CAA = cAA %>% filter(month %in% 2)
m6.CAA = cAA %>% filter(month %in% 6)
m7.CAA = cAA %>% filter(month %in% 7)
m8.CAA = cAA %>% filter(month %in% 8)

# average arr_delay for each month
CAA.M12 = mean(m12.CAA[['arr_delay']])
CAA.M1 = mean(m1.CAA[['arr_delay']])
CAA.M2 = mean(m2.CAA[['arr_delay']])
CAA.M6 = mean(m6.CAA[['arr_delay']])
CAA.M7 = mean(m7.CAA[['arr_delay']])
CAA.M8 = mean(m8.CAA[['arr_delay']])

# df CAA
col2 = c(CAA.M12, CAA.M1, CAA.M2, CAA.M6, CAA.M7, CAA.M8)
col1 = c(12, 1, 2, 6, 7, 8)
mad_CAA = as.data.frame(col1)
mad_CAA['carrier'] = 'AA'
mad_CAA['mad_CAA'] = col2
names(mad_CAA)[names(mad_CAA) == "col1"] <- "month"

# filter months for carrier CUA
m12.CUA = cUA %>% filter(month %in% 12)
m1.CUA = cUA %>% filter(month %in% 1)
m2.CUA = cUA %>% filter(month %in% 2)
m6.CUA = cUA %>% filter(month %in% 6)
m7.CUA = cUA %>% filter(month %in% 7)
m8.CUA = cUA %>% filter(month %in% 8)

# average arr_delay for each month
CUA.M12 = mean(m12.CUA[['arr_delay']])
CUA.M1 = mean(m1.CUA[['arr_delay']])
CUA.M2 = mean(m2.CUA[['arr_delay']])
CUA.M6 = mean(m6.CUA[['arr_delay']])
CUA.M7 = mean(m7.CUA[['arr_delay']])
CUA.M8 = mean(m8.CUA[['arr_delay']])

# df CUA
col3 = c(CUA.M12, CUA.M1, CUA.M2, CUA.M6, CUA.M7, CUA.M8)
mad_CUA = as.data.frame(col1)
mad_CUA['carrier'] = 'UA'
mad_CUA['mad_CUA'] = col3
names(mad_CUA)[names(mad_CUA) == "col1"] <- "month"

# filter months for carrier CDL
m12.CDL = cDL %>% filter(month %in% 12)
m1.CDL = cDL %>% filter(month %in% 1)
m2.CDL = cDL %>% filter(month %in% 2)
m6.CDL = cDL %>% filter(month %in% 6)
m7.CDL = cDL %>% filter(month %in% 7)
m8.CDL = cDL %>% filter(month %in% 8)

# average arr_delay for each month
CDL.M12 = mean(m12.CDL[['arr_delay']])
CDL.M1 = mean(m1.CDL[['arr_delay']])
CDL.M2 = mean(m2.CDL[['arr_delay']])
CDL.M6 = mean(m6.CDL[['arr_delay']])
CDL.M7 = mean(m7.CDL[['arr_delay']])
CDL.M8 = mean(m8.CDL[['arr_delay']])

# df CDL
col4 = c(CDL.M12, CDL.M1, CDL.M2, CDL.M6, CDL.M7, CDL.M8)
mad_CDL = as.data.frame(col1)
mad_CDL['carrier'] = 'DL'
mad_CDL['mad_CAA'] = col4
names(mad_CDL)[names(mad_CDL) == "col1"] <- "month"

# find the mean average of distance for each carrier for each month

# average distance for each month
CAA.M12.d = mean(m12.CAA[['distance']])
CAA.M1.d = mean(m1.CAA[['distance']])
CAA.M2.d = mean(m2.CAA[['distance']])
CAA.M6.d = mean(m6.CAA[['distance']])
CAA.M7.d = mean(m7.CAA[['distance']])
CAA.M8.d = mean(m8.CAA[['distance']])

# df CAA
col5 = c(CAA.M12.d, CAA.M1.d, CAA.M2.d, CAA.M6.d, CAA.M7.d, CAA.M8.d)
md_CAA = as.data.frame(col1)
md_CAA['carrier'] = 'AA'
md_CAA['md_CAA'] = col5
names(md_CAA)[names(md_CAA) == "col1"] <- "month"

# average distance for each month
CUA.M12.d = mean(m12.CUA[['distance']])
CUA.M1.d = mean(m1.CUA[['distance']])
CUA.M2.d = mean(m2.CUA[['distance']])
CUA.M6.d = mean(m6.CUA[['distance']])
CUA.M7.d = mean(m7.CUA[['distance']])
CUA.M8.d = mean(m8.CUA[['distance']])

# df CUA
col6 = c(CUA.M12.d, CUA.M1.d, CUA.M2.d, CUA.M6.d, CUA.M7.d, CUA.M8.d)
md_CUA = as.data.frame(col1)
md_CUA['carrier'] = 'UA'
md_CUA['md_CUA'] = col6
names(md_CUA)[names(md_CUA) == "col1"] <- "month"

# average distance for each month
CDL.M12.d = mean(m12.CDL[['distance']])
CDL.M1.d = mean(m1.CDL[['distance']])
CDL.M2.d = mean(m2.CDL[['distance']])
CDL.M6.d = mean(m6.CDL[['distance']])
CDL.M7.d = mean(m7.CDL[['distance']])
CDL.M8.d = mean(m8.CDL[['distance']])

# df CUA
col7 = c(CDL.M12.d, CDL.M1.d, CDL.M2.d, CDL.M6.d, CDL.M7.d, CDL.M8.d)
md_CDL = as.data.frame(col1)
md_CDL['carrier'] = 'DL'
md_CDL['md_CDL'] = col7
names(md_CDL)[names(md_CDL) == "col1"] <- "month"

# add col to row (month, carrier, mad)
names(mad_CAA)<-c("month","carrier", "mean_arr_delay")
names(mad_CUA)<-c("month","carrier", "mean_arr_delay")
names(mad_CDL)<-c("month","carrier", "mean_arr_delay")

mean_arr_delay <- rbind(mad_CAA, mad_CUA, mad_CDL)

# add col to row (month, carrier, md)
names(md_CAA)<-c("month","carrier", "mean_distance")
names(md_CUA)<-c("month","carrier", "mean_distance")
names(md_CDL)<-c("month","carrier", "mean_distance")

mean_distance <- rbind(md_CAA, md_CUA, md_CDL)

# combine two dataframes for plotting
df.gg = merge(mean_arr_delay, mean_distance)

# Plot `month` against `mean_arr_delay` with `shape` designated by `carrier` and `color` by `mean_distance`
p1 = ggplot(df.gg, aes(x = month, y = mean_arr_delay)) + theme_bw() + geom_point(aes(colour = mean_distance, shape = carrier)) 

# Plot `month` against `mean_arr_delay` with `shape` designated by `carrier` and `color` by `mean_distance` and annotate each point by its associated `carrier` name.
p1 + annotate("text", x = 1, y = 1.1580524, label = "AA") + annotate("text", x = 12, y = 7.2236842, label = "AA") + annotate("text", x = 2, y = 0.9646570, label = "AA") + annotate("text", x = 6, y = 6.3943876, label = "AA") + annotate("text", x = 7, y = 4.1105874, label = "AA") + annotate("text", x = 8, y = -2.4747438, label = "AA") + annotate("text", x = 1, y = -4.0012147, label = "UA") + annotate("text", x = 12, y = 6.1796728, label = "UA") + annotate("text", x = 2, y = -4.4935567, label = "UA") + annotate("text", x = 6, y = 13.7785438, label = "UA") + annotate("text", x = 7, y = 15.1189711, label = "UA") + annotate("text", x = 8, y = 1.3104541, label = "UA") + annotate("text", x = 1, y = 3.6774970, label = "DL") + annotate("text", x = 12, y = 13.7297178, label = "DL") + annotate("text", x = 2, y = 0.4494093, label = "DL") + annotate("text", x = 6, y = 12.0054801, label = "DL") + annotate("text", x = 7, y = 9.9735931, label = "DL") + annotate("text", x = 8, y = 3.6012848, label = "DL")
```

\newpage

# Problem 2

Please refer to the data set `mpg` that is available from the `ggplot2` package. Plot `displ` against `hwy` with faceting by `drv` and `cyl`, `color` disgnated by `class`, and `shape` by `trans`. This illustrates visualization with 4 factors.

```{r, p2, echo = TRUE, warning = FALSE}
# load data
data(mpg)

# plot `displ` against `hwy` 
pa = ggplot(mpg, aes(x = displ, y = hwy)) 

# with faceting by `drv` and `cyl`
pb = pa + facet_grid(drv ~ cyl)

# `color` disgnated by `class`, and `shape` by `trans`
pc = pb + theme_bw() + geom_point(aes(colour = class, shape = trans)) 

# final plot
pc
```